---
title: "Plotting external data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting external data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(eval = TRUE, fig.retina = 3, fig.width = 6)
```

Most of the time, you're not plotting empty atlases. You have results --
p-values, cortical thickness, whatever -- and you want them on a brain. This
vignette covers how to get your data into the right shape for ggseg.

```{r}
#| label: load-packages
library(ggseg)
library(dplyr)
library(ggplot2)
```

## How matching works

`geom_brain()` joins your data to the atlas by any columns they share.
That means your data needs at least one column with names that match the
atlas. The two columns you'll use most:

- **region** -- human-readable names like "insula" or "precentral"
- **label** -- FreeSurfer labels like "lh_bankssts"

Check what's available:

```{r}
#| label: dk-regions
ggseg.formats::atlas_regions(dk())
```

```{r}
#| label: dk-labels
ggseg.formats::atlas_labels(dk())
```

Names must match exactly, including case and spacing.

## A minimal example

Three regions, three p-values:

```{r}
#| label: minimal-data
some_data <- tibble(
  region = c("superior temporal", "precentral", "lateral orbitofrontal"),
  p = c(.03, .6, .05)
)
some_data
```

Pass the data to `ggplot()` and map `fill` to your variable:

```{r}
#| label: fig-minimal-plot
#| fig-cap: "Brain plot with three regions coloured by p-value."
ggplot(some_data) +
  geom_brain(atlas = dk(), mapping = aes(fill = p))
```

Regions not in your data appear as `NA` (grey by default). Regions in your
data that don't match the atlas are silently dropped, so watch your spelling.

## Constraining matches with extra columns

If your data is hemisphere-specific, add a `hemi` column. The join will use
both `region` and `hemi`, so values only land on the correct side:

```{r}
#| label: fig-hemi-constraint
#| fig-cap: "Brain plot restricted to the left hemisphere using a hemi column."
some_data$hemi <- "left"

ggplot(some_data) +
  geom_brain(atlas = dk(), mapping = aes(fill = p))
```

The same works for any atlas column -- adding `view`, for instance, would
restrict matches to specific views.

## Faceting across groups

If your data has a grouping variable, `facet_wrap()` and `facet_grid()` work
as you'd expect. `geom_brain()` detects the faceting variables and
replicates the full atlas in each panel:

```{r}
#| label: fig-facet-groups
#| fig-cap: "Brain plots faceted by age group with a custom colour gradient."
some_data <- tibble(
  region = rep(
    c(
      "transverse temporal",
      "insula",
      "precentral",
      "superior parietal"
    ),
    2
  ),
  p = sample(seq(0, .5, .001), 8),
  group = c(rep("Young", 4), rep("Old", 4))
)

ggplot(some_data) +
  geom_brain(atlas = dk(), colour = "white", mapping = aes(fill = p)) +
  facet_wrap(~group, ncol = 1) +
  theme(legend.position = "bottom") +
  scale_fill_gradientn(
    colours = c("royalblue", "firebrick", "goldenrod"),
    na.value = "grey"
  )
```

No need to call `group_by()` first -- the geom handles atlas replication
automatically. (Explicit `group_by()` still works for backward
compatibility.)

## The pre-merged workflow

For full control over faceting or when you need to combine brain data with
other sf layers, convert the atlas to a data frame and join manually:

```{r}
#| label: atlas-columns
atlas_df <- as.data.frame(dk())
names(atlas_df)
```

Then use a standard join and `geom_sf()`:

```{r}
#| label: fig-pre-merged
#| fig-cap: "Brain plot using a manual left_join and geom_sf for full control."
some_data <- tibble(
  region = c("superior temporal", "precentral", "lateral orbitofrontal"),
  p = c(.03, .6, .05)
)

atlas_df |>
  left_join(some_data, by = "region") |>
  ggplot() +
  geom_sf(aes(fill = p), colour = "white") +
  facet_grid(hemi ~ view) +
  theme_void()
```

See `vignette("geom-sf")` for more on this approach.

## Troubleshooting

**Regions don't show up.** Check spelling and case. `ggseg.formats::atlas_regions(dk())`
gives you the exact strings the atlas expects.

**Data lands on both hemispheres.** Add a `hemi` column with `"left"` or
`"right"` to constrain the match.

**Extra facet panels appear.** This is handled automatically by
`geom_brain()`. If you're using the `brain_join()` + `geom_sf()` workflow
directly, `group_by()` your data by the faceting variable before joining.
