---
title: "How geom_brain() works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How geom_brain() works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 2.4
)
set.seed(1)
```

The first time you plot three regions and get a whole brain back, it can look like sleight of hand.
You handed `geom_brain()` a few rows; it drew the entire cortex, greyed the regions you left out, and coloured the ones you named.
There is no trick -- just three small ggplot2 pieces doing their jobs.
Knowing what each one does explains why faceting, aggregation, and colour scales behave the way they do, and how to build on them.

```{r}
#| label: load
library(ggseg)
library(ggplot2)
```

## The brain shape is drawn by a stat

The whole brain always renders, even when your data covers only a handful of regions.
That comes from `StatBrain`, a ggplot2 stat that _generates_ the atlas geometry instead of reading it from your data.
It is the same move `geom_sf()` makes: the shape is computed, not supplied.

Map nothing and there is nothing of yours to colour by, so `geom_brain()` falls back to the atlas's own palette: it maps `fill` to `label`, the key the palette is built on.
That is the atlas looked at for its own sake.

```{r}
#| label: bare
#| fig-alt: "The DK atlas with every region in its own atlas palette colour."
#| fig-cap: "A bare atlas: the full shape, in the atlas palette."
ggplot() +
  geom_brain(atlas = dk(), show.legend = FALSE)
```

Give it data and your mapping takes over: the regions you have values for take colour from your scale.
Everything else stays grey, so the brain remains legible even when you have measured only a corner of it.

```{r}
#| label: subset
#| fig-alt: "Three regions on a red-blue diverging scale, the rest grey."
#| fig-cap: "Three regions have values; the rest of the brain still renders."
results <- data.frame(
  region = sort(unique(ggseg.formats::atlas_regions(dk())))[1:3],
  score = c(2.1, -1.4, 0.8)
)

ggplot(results, aes(fill = score)) +
  geom_brain(atlas = dk(), show.legend = FALSE) +
  scale_fill_gradient2()
```

## Repeated regions are combined

Real data rarely arrives one row per region.
You might have one row per subject, or per session, or per run.
When several rows land on the same region, `StatBrain` reduces them to a single value before drawing.
The default is the mean, and the `fun` argument takes any function that turns a vector into one number.

```{r}
#| label: aggregate
#| fig-alt: "The three regions summarised by mean, then by maximum."
#| fig-cap: "Repeated-measures data summarised by mean and by maximum."
thickness <- data.frame(
  region = rep(sort(unique(ggseg.formats::atlas_regions(dk())))[1:3], each = 4),
  subject = rep(1:4, times = 3),
  thickness = c(2.5, 2.6, 2.4, 2.7, 3.1, 3.0, 3.2, 2.9, 2.8, 2.7, 2.9, 2.6)
)

# mean per region (the default)
ggplot(thickness, aes(fill = thickness)) +
  geom_brain(atlas = dk(), show.legend = FALSE)

# the maximum instead
ggplot(thickness, aes(fill = thickness)) +
  geom_brain(atlas = dk(), fun = max, show.legend = FALSE)
```

## Faceting comes for free

Here is where being a real stat pays off.
ggplot2 recomputes a stat once per facet panel, so the complete brain appears in every panel straight from your data.
No `group_by()`, no copying the atlas by hand -- `facet_wrap()` and `facet_grid()` behave exactly as they do for any other geom.

```{r}
#| label: facet
#| fig-height: 4
#| fig-alt: "Two stacked brains, one per cohort, four regions coloured."
#| fig-cap: "One brain per cohort, each drawn from that cohort's own rows."
cohorts <- expand.grid(
  region = sort(unique(ggseg.formats::atlas_regions(dk())))[1:4],
  group = c("patients", "controls"),
  stringsAsFactors = FALSE
)
cohorts$score <- rnorm(nrow(cohorts))

ggplot(cohorts, aes(fill = score)) +
  geom_brain(atlas = dk(), show.legend = FALSE) +
  facet_wrap(~group, ncol = 1) +
  scale_fill_gradient2()
```

Each panel is summarised on its own rows, so the same region can read differently from one panel to the next.

## Colour, including bins, is just a scale

`StatBrain` hands the geom an ordinary continuous `fill`.
That one fact means every ggplot2 fill scale works without ggseg doing anything special.

So how would you make a binned map -- discrete steps instead of a smooth gradient?
You do not reach for a ggseg feature; you reach for a scale.
A stepped scale bins the aggregated value for you.

```{r}
#| label: bins-scale
#| fig-alt: "The atlas in four discrete colour steps from a binned scale."
#| fig-cap: "Binning the aggregated value with a stepped scale."
ggplot(thickness, aes(fill = thickness)) +
  geom_brain(atlas = dk()) +
  scale_fill_steps(n.breaks = 4)
```

When your bins carry meaning -- a threshold, a clinical band -- cut the values yourself and map the resulting factor.

```{r}
#| label: bins-cut
#| fig-alt: "Regions coloured by a factor: thin, typical, and thick."
#| fig-cap: "Named bands, made with cut() and mapped as a discrete fill."
thickness$band <- cut(
  thickness$thickness,
  breaks = c(-Inf, 2.7, 3.0, Inf),
  labels = c("thin", "typical", "thick")
)

ggplot(thickness, aes(fill = band)) +
  geom_brain(atlas = dk())
```

The division of labour is the point: the stat decides how many rows become one value, and the scale decides how that value becomes a colour.
Continuous gradients, viridis, diverging, binned, stepped, and manual palettes all compose the same way.

## The three pieces, if you want to build on them

Under `geom_brain()` sit three ggproto objects, each with a narrow job.

`StatBrain` generates the atlas geometry per panel, reduces your values per region with `fun`, and joins them onto the shape.
It emits the polygon `x` and `y`, the `group` and `subgroup` that encode ring structure and holes, and passes your `fill` straight through.

`GeomBrain` is a thin `GeomPolygon` subclass.
Its only additions are the brain defaults -- a `grey35` outline and a grey fill for an atlas with no palette of its own -- and both step aside the moment you map an aesthetic of your own.

`LayerBrain` is the layer that runs at plot-build time.
It flattens the atlas, hands the geometry to the stat, and quietly maps the join keys as aesthetics so `label` and `region` survive into the stat.
Running at build time is also what lets it see a `fill` set in the top-level `ggplot()` call before deciding whether to supply the palette default.

`stat_brain()` is the stat-first spelling of `geom_brain()` -- both build the same layer.
Reach for it when you want `StatBrain` paired with a different geom, such as drawing the atlas vertices as a point cloud.

```{r}
#| label: stat-brain
#| fig-alt: "The atlas as a dense cloud of small grey points."
#| fig-cap: "StatBrain with a point geom instead of the polygon default."
ggplot() +
  stat_brain(atlas = dk(), geom = "point", size = 0.1, show.legend = FALSE)
```

Everything else you already use -- `position_brain()` layouts, `coord_brain()`, the `scale_*_brain()` palettes, `theme_brain()` -- sits on top of these three pieces without changes.
